A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-16 model.tex
improved pdv interpret

\subsubsection{Physical picture of dust-drag instabilities}

The full energy equation is 
\begin{align*}
  \frac{DP}{Dt} = \frac{P}{\rho}\frac{D\rho}{D t} + \mathcal{C}. 
\end{align*}
Dust-drag causes a phase difference between the 
pressure and density evolution of a fluid element. Now, consider a 
fluid element at maximum density during an oscillation cycle, 
so that $D\rho/ D t = 0$. If at this point dust-gas drag provides an
effective heating to the fluid element ($\mathcal{C}>0$, i.e. a particle
flux \emph{away} from it), then it will 
experience increasing pressure, $DP/D t>0$. It will attain  
pressure maximum \emph{after} reaching density maximum. This implies 
pressure maximizes while the parcel is expanding: it will do 
positive work  --- the area enclosed by the parcel's path in the `$P$-$V$' plane  is 
positive. This will increase the oscillation amplitude.

@@

We suggest understanding dusty instabilities in
protoplanetary disks by re-casting the problem into that of
thermodynamics. Then dust-drag instabilities appear in the equations in the 
same way as pulsational instabities in stars, which has already been 
well studied. Whether or not the dusty disk is actually unstable
depends on details of the dust-drag or `cooling' function
$\mathcal{C}$. However, given the complexity of $\mathcal{C}$ (see
Appendix \ref{lin_dust}), we might generally expect the system to 
support a range of stable and unstable modes. On physical grounds,
growing modes \emph{must} be so by causing a phase lag between
pressure and total density.

>>>>

\subsubsection{Physical picture of dust-drag instabilities} 
The full energy equation is 
\begin{align*}
  \frac{DP}{Dt} = \frac{P}{\rho}\frac{D\rho}{D t} + \mathcal{C}. 
\end{align*}
Dust-drag causes a phase difference between the pressure and density 
evolution of a fluid element. In Fig. \ref{pdv_cartoon}, we show that
if pressure lags behind density, then positive
work done ($\mathcal{W}>0$) and would lead to growing oscillations
($s>0$).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of overstable modes caused by
    dust-gas drag. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the $P$-$V$ plane. In the case shown,
    dust-drag causes 
    pressure to lag behind density, which results in a clockwise path in
    the $P$-$V$ plane, implying positive work done by the fluid and
    hence would lead to growing oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

Consider a fluid element at maximum density during an oscillation cycle
(point $A$), so that $D\rho/ D t = 0$. If dust-gas drag provides an
effective heating to the fluid element at $A$ (i.e. gas influx)  then
it will experience increasing pressure, $DP/D t>0$. It attains
pressure maximum (point $B$) \emph{after} reaching density maximum.   
This implies pressure maximizes while the parcel is expanding: 
the parcel's path in the `$P$-$V$' plane  is clockwise, as shown in
the bottom panel of Fig. \ref{pdv_cartoon}. This corresponds to
positive work done.

Our thermodynamic interpretation does not rigorously explain
\emph{why} dust drag causes a pressure lag behind density, but merely shows
that this \emph{must} be the case for any growing modes associated the
dust drag. To understand how dust-drag causes this lag requires an
explicit solution to the linearized equations with detailed treatment
of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
the system to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

@@
